Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-02-28 MCMC-appendix.tex
write most of intro

\begin{document}
\ifSubfilesClassLoaded{
  \setcounter{chapter}{2}
}

\chapter{Statistical and epidemiological background} \label{inc-prev}

\todo[inline]{%
  Chapter 3 needs a better title: this is not all the statistical and epidemiological background and some of it is mathematical modelling / disease dynamics.
  Perhaps: ``Data generating procsses and disease dynamics''.
  Would need to define disease dynamics.
}

%\todo[inline]{This chapter is pretty short. Not sure if this matters or it needs to merged with another chapter? For example, one background chapter split into epi/biology and stats.}

This chapter defines incidence and prevalence, and the relationship between these quantities.
In the course of explaining this relationship, it will become clear that the duration of RT-PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.

I start by defining these terms in \cref{inc-prev:sec:definitions}.
I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.

The infection process (\cref{inc-prev:sec:infection-process}) models how infections arise.
The concepts here are well-established within infectious disease epidemiology (see\todo{ref intro chapter}).

The prevalence process (\cref{inc-prev:sec:prevalence-process}) describes the relationship between incidence and prevalence.
Here, I adapt previous work to the SARS-CoV-2 context.
In particular, recovering from infection needs to be considering, which was not the case for HIV/AIDS.

The observation process (\cref{inc-prev:sec:observation-process}) describes the relationship between prevalence and the observed data, ultimately allowing inference of incidence based on this data.
Unlike in HIV/AIDS backcalculation, the CIS has a sampling mechanism that needs to be modelled.
This model is guided by the specific study design of the CIS.

Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis.

\Cref{inc-prev:sec:conclusion} concludes the chapter.

\section{Definitions} \label{inc-prev:sec:definitions}

Time will be considered discrete in this chapter, except in \cref{SEIR:sec:mechanistic-models}.
I use days, however, the results generalise in the obvious way to other time units.
I choose discrete time for pragmatic reasons.
Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.

\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
As discussed in\todo{ref intro}, the timing of infection events is very rarely directly observable and must be inferred from other data.
Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.

\emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
I refer to individuals with the disease as \emph{prevalent} individuals.
Whether an individual is prevalent is often expensive or error-prone to measure, therefore, it will often be regarded as a latent quantity.
Furthermore, what is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
Within this thesis, I define an individual as prevalent on day $t$ if they are detectable (detectable is defined in \cref{E-biology-data:sec:detectable}).
I assume that an individual's prevalent status cannot change over the course of a day.
Therefore, prevalence is constant over any single day $t$.
I denote the number of prevalent individuals on day $t$ as $P_t$.
Therefore, the prevalence is $P_t/\Npop$.

\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
$D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
I model the $D_i$s as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be expressed in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.

Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
This implies that $F_D(d) \in [0, 1]$ and increases monotonically.
Therefore, $S(d) \in [0, 1]$ and decreases monotonically.
Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterisation for inference.

\section{Infection process} \label{inc-prev:sec:infection-process}
The infection process is the mathematical description of how infections occur.
It is a \emph{point} or \emph{counting} process, where each infection is an event.

A point process is a random function of continuous time $K(t)$.
$K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocite[244]{yanDistribution}.
In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
I will explain two approaches to modelling the infection process.

First, in \cref{inc-prev:sec:poisson-process}, as a \emph{Poisson process}.
A Poisson process is a widely-used type of counting process both within and outside of epidemiology.
It makes no direct assumptions on the underlying biological or epidemiological process, and is therefore a flexible model.

Afterwards, in \cref{SEIR:sec:mechanistic-models}, I introduce \emph{mechanistic models}.
Mechanistic models explicitly describe the population and transmission between individuals.
They vary greatly in their details and realism~\autocite{murilloMultiscale}.

\subsection{Poisson processes} \label{inc-prev:sec:poisson-process}

A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod,paganoHIV,rosenbergBackcalculation,brookmeyerBackcalculation}.
% A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
Extensions to this model, which I do not consider here, have incorporated covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.

The first defining property of a Poisson process is that the counts are Poisson distributed.
Formally, $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the infection process at time $r$~\autocite[244]{yanDistribution}.

The other defining property of a Poisson process is that it has \emph{independent increments}.
A counting process has independent increments if the number of events in disjoint intervals are independent conditional on the intensity~\autocite[244]{yanDistribution}.
That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
In the context of a Poisson process, the independent increment property implies that, conditional on $\nu(r)$, the number of infections in disjoint intervals are independent.
Hence, $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.

The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic~\autocite{beckerDependent,brookmeyerMethod}.

Two features of a Poisson process are violated for epidemics.
Epidemics commonly exhibit \emph{overdispersion} and have feedback loops~\autocite{beckerDependent}.

Overdispersion means that $\V(Z_t) > \E(Z_t)$.
In a Poisson process $Z_t$ has a Poisson distribution, hence, $\V(Z_t) = \E(Z_t)$.
Therefore, an overdispersed epidemic modelled as a Poisson process will be less variable than the realised epidemic.
Overdispersion is common in infectious diseases, especially for coronaviruses such as SARS-CoV-2~\autocite{endoEstimating,adamClustering,mccloskeySARS}.

% A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
% A super-spreading event is an event where many more people than average are infected.
% That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
% Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocite{endoEstimating,adamClustering}{mccloskeySARS}.
% An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
% One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
% For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
% Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
% Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
% %The negative binomial tends to Poisson as $k\to\infty$.
% Generally, $k<1$ is considered as significant levels of overdispersion.
% Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocite{quiltyReconstructing,quiltyUnderstanding}.
% If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
% If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
% A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
% The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.

Feedback loops mean that stochastically low or high numbers of infections affect the future number of infections.
That is, $\nu(t+\Delta t)$, for $\Delta t > 0$, is dependent on $K(t)$, even after conditioning on $\nu(t)$.
This violates the assumption of independent increments.
Infectious diseases generate feedback loops because stochastically high or low incidence feeds back to the process intensity at future times.
% This can be seen most obviously in the event of a super-spreading event, when many .
% If a super-spreading event occurs, then many more individuals than suggested by the process intensity will be infected.
% These additional infections will propagate the epidemic further, increasing the intensity at future times.
Point processes can be generalised to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.

Both overdispersion and feedback loops violate the properties of a Poisson process.
Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.

% In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/\Npop$.
% However, this is often a poor approximation.
% \emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).

% Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
% The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
% I keep this assumption for the remainder of this chapter.

% Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.

% Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
% This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.

\subsection{Mechanistic models} \label{SEIR:sec:mechanistic-models}

Now I introduce mechanistic models, in particular the class of mechanistic transmission models known as \emph{compartmental} models.
Compartmental models are the most widely-used class of mechanistic infectious disease model.
A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
I will describe these models using the notation of the field, which differs in conventions from the rest of this thesis.
In particular, upper case or lower case will no longer be used to distinguish between random variables and their realisations, which instead is clear by context.

A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$~\autocite{birrellEvidence,dukicTracking,corbellaThesis}.
The elements of $\vec{x}(t)$ are the latent proportion of the population (or number of individuals) in each compartment at time $t$.
The parameters $\vec{\theta}$ determine, along with the model structure, how the state vector changes over time.
% Inference about $\vec{x}(t)$ and $\vec{\theta}$ is requires linking to data, $\vec{y}$, via an observation model, $p(\vec{y}(t) \mid \vec{x}, \vec{\theta})$ where $\vec{x}$ denotes conditioning on $\vec{x}(t)$ for all $t$~\autocite{birrellEvidence} (see \cref{SEIR:sec:observation}).
% That is, the observation model specifies the likelihood of the data given the model state.

Compartmental models can be \emph{deterministic} or \emph{stochastic}.
Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \vec{x}(0), \theta) = 0$~\autocite{birrellEvidence}.
% In most situations, the function is not analytically-tractable and requires numeric methods to solve.
%The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
%There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).

Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters~\autocite{birrellEvidence,dukicTracking,corbellaThesis,keelingModeling}.
Stochastic models are most commonly specified in discrete time, using a Markov assumption.
A common simplification is to use discrete time and a Markov assumption; \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).

Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of a corresponding stochastic model.
When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
This is especially true when making noisy observations of the system, and that the stochastic differences between the stochastic and its mean are negligible compared to the observation noise.
I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.

The remainder of this section uses only deterministic models.
This choice is appropriate for the context of this thesis because there were many individuals infectious with SARS-CoV-2 in this time period.
Hence, further details on stochastic models are not outside the scope this thesis; a wealth of introductory texts with further details on stochastic models are available, I recommend \textcite[chapter 6]{keelingModeling}.

% This section discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
% In an ODE model, time is necessarily continuous.%, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
% For concreteness, I use units of days throughout, although the model can use any unit of time.
% The number or proportion of individuals in each compartment is also continuous; clearly, this is not true in reality but is a good approximation when the population is large.

% In this section, I introduce the \emph{susceptible-infectious-recovered} (SIR) model and the extensions to it I require to appropriately model SARS-CoV-2.

% \Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infectious-recovered} (SIR) model, the simplest compartmental model for a disease where individuals are immune to reinfection following their recovery.
% The name SIR derives from the structure of the model.
% In particular, the model divides the population into one of three states: susceptible, infectious, or recovered.

% This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, in the remainder of this section I introduce standard extensions to the SIR model which relax these assumptions.
% First, in \cref{SEIR:sec:non-exponential}, I relax the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
% Then, in \cref{SEIR:sec:SEIR}, I relax the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
% Finally, in \cref{SEIR:sec:structured-populations}, I include heterogeneity in the population, stratifying by important characteristic(s).
% Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.

\subsubsection{The SIR model} \label{SEIR:sec:SIR}

\begin{figure}[h]
\makebox[\textwidth][c]{
\begin{tikzpicture}[
    node distance = 2.5cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle},
    ]

    \node[state] (S) {$s$};
    \node[state, right=of S] (I) {$i$};
    \node[state, right=of I] (R) {$r$};

    \path (S) edge node {$\lambda(t)$} (I)
          (I) edge node {$\gamma$} (R);
\end{tikzpicture}
}
  \caption[The SIR model]{Schematic of the basic SIR model. Arrows are labelled with the rates at which individuals move between compartments.}
  \label{SEIR:fig:SIR}
\end{figure}

The simplest compartmental model for a disease which confers immunity is the SIR (susceptible-infectious-recovered) model~\autocite[15]{keelingModeling}, depicted in \cref{SEIR:fig:SIR}.
The name SIR derives from the structure of the model.
In particular, the model divides the population (assumed closed, an assumption discussed later) into one of three compartments: susceptible, infectious, or recovered.
% The state of the system is the proportion of the population in each compartment.
In the SIR model, individuals are assumed to be homogeneous except for their disease state, \ie they are otherwise identical.

Individuals in the susceptible compartment are those that could be infected (they have no immunity).
The proportion of the population that is susceptible at time $t$ is $s(t)$.
Individuals in the infectious compartment can infect others, and eventually recover.
The proportion of the population that is infectious at time $t$ is $i(t)$.
Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
The proportion of the population that have recovered by time $t$ is $r(t)$.
The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.

The SIR model is described by the following system of ODEs (ordinary differential equations)~\autocite[19]{keelingModeling}:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t) \\
\frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
\frac{dr(t)}{dt} &= \gamma i(t)
\end{align}
where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
The dynamics of the model are described by the parameters $\vec{\theta} = (\beta, \gamma)^T$.
$\beta$ is the transmission rate, absorbing several terms as described below.
$\gamma$ is the recovery rate, or the rate that infectious individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
These parameters are assumed to be constant.

% The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
% Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infectious individual~\autocite[214]{kretzschmarMathematical}.
Assume this individual comes into contact with other individuals in the population at a mean rate of $\kappa$.
Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Therefore the infectious individual makes contact with susceptible individuals at a mean rate of $\kappa s(t)$ per day.
Assume that there is a constant probability of transmission upon contact between an infectious and susceptible individual, $q$.
Hence, the mean rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
The two constant terms are absorbed into the transmission rate, $\beta = \kappa q$.
Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$ per day.
Multiplying by the number of infectious individuals and dividing by the population size gives the rate per day at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.

% The term $\gamma i(t)$, the rate at which infectious individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item For any short period of time when the change in $s(t)$ is negligible, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
    A notable example is the early part of an epidemic when $s(t) \approx 1$.
    This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
    Intuitively, this can be thought of as the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
    $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
    \item The population is assumed to be closed.
    This assumption implies several attributes.
    First, the total population size is constant.
    Second, only a negligible number of infections are imported from outside the population.
    Third, births and deaths of individuals are negligble over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
    \item As noted previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ are required to solve the system.
    \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
    \item The incidence proportion on day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$. However, this will not in general give an integer value to $Z_t$. In practice, this is not an issue since a continuous approximation for the incidence will suffice.
\end{itemize}

\subsubsection{Non-exponential waiting times} \label{SEIR:sec:non-exponential}
\begin{figure}[h]
\makebox[\textwidth][c]{
\begin{tikzpicture}[
    node distance = 2.5cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle},
    ]

    \node[state] (S) {$s$};
    \node[state, right=of S] (I1) {$i_1$};
    \node[state, right=of I1] (I2) {$i_2$};
    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
    \node[state, right=of I3] (I4) {$i_n$};
    \node[state, right=of I4] (R) {$r$};

    \path (S) edge node {$\beta i$} (I1)
          (I1) edge node {$n\gamma$} (I2)
          (I2) edge node {$n\gamma$} (I3)
          (I3) edge node {$n\gamma$} (I4)
          (I4) edge node {$n\gamma$} (R);
\end{tikzpicture}
}
  \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having multiple I states.}
  \label{SEIR:fig:SIR-gamma}
\end{figure}

The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
The ODEs are now:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t)
\end{align}
where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.

The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$.

\subsubsection{The SEIR model} \label{SEIR:sec:SEIR}

\todo[inline]{Maybe merge SEIR intro into the general intro, not introducing SIR first}

\begin{figure}[h]
\makebox[\textwidth][c]{
\begin{tikzpicture}[
    node distance = 2cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle},
    % remember picture, overlay
    ]

    \node[state, xshift=-2cm] (S) {$s$};
    \node[state, right=of S] (E1) {$e_1$};
    \node[state, right=of E1] (E2) {$e_2$};
    \node[state, right=of E2, draw=none] (E3) {$\cdots$};
    \node[state, right=of E3] (E4) {$e_m$};
    % \node[state, right=of E4] (I1) {$i_1$};
    \node[state, below=of E1] (I1) {$i_1$};
    \node[state, right=of I1] (I2) {$i_2$};
    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
    \node[state, right=of I3] (I4) {$i_n$};
    \node[state, right=of I4] (R) {$r$};

    \path (S) edge node {$\beta i$} (E1)
          (E1) edge node {$m\sigma$} (E2)
          (E2) edge node {$m\sigma$} (E3)
          (E3) edge node {$m\sigma$} (E4)
          (I1) edge node {$n\gamma$} (I2)
          (I2) edge node {$n\gamma$} (I3)
          (I3) edge node {$n\gamma$} (I4)
          (I4) edge node {$n\gamma$} (R);
    \draw[->] (E4.east) -- ++(1,0) |- ([yshift=-1cm]E4.east) node [near start, right] {$m\sigma$} -|  (I1.north);
\end{tikzpicture}
}
  \caption[The SEIR model]{Schematic of the SEIR model.}
  \label{SEIR:fig:SEIR}
\end{figure}

The SIR model assumes that individuals are immediately infectious upon infection, neglecting the latent period (see \cref{E-biology-data:sec:natural-history}).
Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
The SIR model can be extended by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.

Similar to the infectious period, the latent period can be modelled as a gamma distribution using multiple latent compartments, $e_1, dots, e_m$.
\begin{align}
\frac{ds}{dt} &= -\beta si \\
\frac{de_1}{dt} &= \beta si - m\sigma e_1 \\
\frac{de_2}{dt} &= m\sigma e_1 - m \sigma e_2 \\
&\vdots \nonumber \\
\frac{de_m}{dt} &= m\sigma e_{m-1} - m \sigma e_m \\
\frac{di_1}{dt} &= m\sigma e_m - n\gamma i_1 \\
\frac{di_2}{dt} &= n\gamma i_1 - n \gamma i_2 \\
&\vdots \nonumber \\
\frac{di_n}{dt} &= n\gamma i_{n-1} - n \gamma i_n \\
\frac{dr}{dt} &= n\gamma i_n
\end{align}
where $1/\sigma$ is the mean duration of the latent period.
The latent period is distributed $\GamDist(m, m\sigma)$ and the infectious period is still distributed $\GamDist(n, n\gamma)$.

The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.

\subsubsection{Structured populations} \label{SEIR:sec:structured-populations}

To relax the assumption of a homogeneous population, the population can be stratified.
If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
I denote the set of strata, as $\set{A}$, with the set's size as $A$.
I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
Stratification requires two major  model changes: extension of the state vector and a description of the between-strata interactions.

The state vector is now a vector of vectors, one for each of the compartments
For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
The $a$th element of each vector ($a \in \set{A}$) is the proportion of strata $a$ in that compartment.
For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
These vectors replace the scalars in the previous equations.
Other, equivalent formulations of this model exist within the literature.
For example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.

The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Following the same argument as in \cref{SEIR:sec:SIR} finds that the infectious individual makes contact with susceptible individuals in stratum $a$ at a mean rate of $\kappa_{aa'} s_a(t)$ per day.
Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
Therefore, the mean rate of infections in strata $a$ per day caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, and dividing by the size of strata $a$, $N_a$, gives $\kappa_{aa'} q_{aa'} s_a(t) \frac{1}{N_a} N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} \frac{1}{N_a} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
This gives the total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.

>>>>

\begin{document}

\chapter{Markov chain Monte Carlo} \label{MCMC}

This appendix outlines the basic theory and implementation of Markov chain Monte Carlo (MCMC) methods.
Any reader interested in more technical details or proofs of the results here should consult the references within4.

\section{Metropolis-Hastings}
